On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-23 paper/model3d.tex
reg update
model version 2

\section{Basic equations}\label{basic}
We consider a three-dimensional (3D), self-gravitating,
viscous disk with heating and cooling. We use the shearing box framework to
study a small patch of the disk \citep{goldreich65}. The local frame
co-rotates with a fiducial point in the unperturbed disk at angular
frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ corresponds
to the radial, azimuthal and verical directions in the global disk. 
The fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) = 0, \\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi - 2\Omega\hat{\bm{z}}\times\bm{v}\notag\\ & 
  \phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =}
  +2\Omega^2qx\hat{\bm{x}} -\Omega_z^2z\hat{\bm{z}} 
   + \frac{1}{\rho}\nabla\cdot\bm{T},\label{momentum_eq}\\
  & \frac{\p E}{\p t} + \nabla\cdot\left(E\bm{v}\right) = - P
  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} +
  \mathcal{H}_\mathrm{ext} - \Lambda,\label{energy_eq}
\end{align} 
where $\rho$ is the density field and $\bm{v} = (v_x, v_y, v_z)$ is
the velocity field.% $q$ is the shear parameter and $\Omega_z$ is the
%local vertical oscillation frequency. 
We assume an ideal gas so that the pressure $P$ and thermal energy
density $E$ are related by  
\begin{align}
  P = (\gamma-1)E = \mathcal{R}\rho T, 
\end{align}
where $\gamma$ is the adiabatic index, $\mathcal{R}$ is the gas
constant, and $T$ is the temperature. We use $\gamma=1.4$ unless
otherwise stated. The gas 
gravitatational potential $\Phi$ is given via the Poisson equation,  
\begin{align}
 \nabla^2\Phi = 4 \pi G \rho, 
\end{align}
where $G$ is the gravitational constant.

@@

In the energy equation (Eq. \ref{energy_eq}) the source terms 
$\mathcal{H}_\mathrm{visc}$, $\mathcal{H}_\mathrm{ext}$ and $\Lambda$
represent viscous heating, any time-independent heat
source/sinks, and time-dependent thermal losses, respectively. These
are described below. We normally set
$\mathcal{H}_\mathrm{ext}=0$ unless otherwise stated.

\subsection{Viscosity and heating}\label{visc_model}
The Cartesian components of the viscous stress tensor $\bm{T}$ is
defined by 
\begin{align}
  T_{ij} \equiv \rho \left[\nu \left(\p_j v_i + \p_i v_j\right) +
    \left(\nu_b-\frac{2}{3}\nu\right)\delta_{ij}\nabla\cdot\bm{v}\right], 
\end{align}
%bulk visc ignored 
where $\rho\nu$ is the shear viscosity, and we include a bulk
viscosity $\rho\nu_b$ for completeness.  
The associated viscous heating given by
\begin{align}
  \mathcal{H}_\mathrm{visc} \equiv \left(\p_jv_i\right)T_{ij}, 
\end{align}
where summation over repeated indices is implied.

We adopt the viscosity law  
\begin{align}
  \nu = \alpha
  \frac{c_{s0}^2}{\Omega}\left(\frac{\rho}{\rho_\mathrm{eq}}\right)^\mu\left(\frac{P}{P_\mathrm{eq}}\right)^\lambda,            
\end{align}
where subscript `eq' denotes the equilibrium state and  
%$c_{s0}$ is a modified sound-speed defined below. 
$c_{s0}^2\equiv P_\mathrm{eq}(0)/\rho_\mathrm{eq}(0)$. 
The dimensionless coefficient
$\alpha=~\alpha(\rho_\mathrm{eq},P_\mathrm{eq})$ characterizes the
magnitude of the shear viscosity \emph{in steady state}. The indices
$\mu,\,\lambda$ are free parameters chosen to model how the viscosity
behaves in the perturbed state.  
%Given the uncertain
%properties of gravito-turbulence, we will treat $\mu,\lambda$ as free
%parameters. 
We adopt the same prescription for the bulk viscosity but
with $\alpha\to\alpha_b$.

Our numerical calculations use $\mu=-1,\lambda=0$ so 
that $\rho\nu$ is time-indpendent, as adopted in previous 
studies of viscous GI 
\citep{lynden-bell74,hunter83,willerding92,gammie96}.  
This choice also conveniently eliminates viscous over-stability
\citep{schmit95,latter06}, which is unrelated to self-gravity, that
would otherwise contaminate our results. %obscure

While steady-state viscosity values can be determined analytically
\citep[e.g.][]{rafikov15}, its time-dependent behavior is not
well-explored. We emphasize the choice $\mu=-1,\lambda=0$ is made to
bring out the physically relevant process --- viscosity-assisted
gravitational instability. %expect to be generally present, but will
                           %be obscured by viscous over-stability with
                           %other indices
Interestingly, though, \cite{laughlin96b} have suggested a $\nu \propto 1/\Sigma$
dependence when modeling the evoluting of self-gravitating disks 
as a viscous process. Furthermore, \cite{paardekooper12} measured
an increase in the effective viscosity when their gravito-turbulent
disk models underwent fragmentation (i.e. as $\Sigma$ increases).

%{\bf note paardekooper 2012: alpha goes down as Sigma goes up}

\subsection{Steady states and cooling models}
We consider equilibrium solutions (here omitting the `eq' subscripts
for clarity) 
\begin{align}
  \bm{v} &= -q\Omega x \hat{\bm{y}}, \\
  \rho   &= \rho(z),\\
  P      &= P(z) \equiv c_s^2(z)\rho.
\end{align} 
%and assume a polytropic relation for the equilibrium pressure,
%\begin{align}
%  P  =
%  \frac{c_{s0}^2\rho_0}{\Gamma}\left(\frac{\rho}{\rho_0}\right)^\Gamma
%  \equiv \frac{c_s^2(z)}{\Gamma}\rho,
%\end{align}
%where $\rho_0$ is the equilibrium midplane density and $\Gamma$ is the
%constant polytropic index.  
The equilibrium density and pressure fields are obtained by solving
the vertical momentum equation with self-gravity,
\begin{align}
  &\frac{1}{\rho}\frac{dP}{dz} +
  \Omega_z^2z + \frac{d\Phi}{dz} = 0, \label{vert_eq1}\\
 &\frac{d^2\Phi}{dz^2} = 4 \pi G \rho,\label{vert_eq2}
\end{align}
%A numerical solution is usually required to obtain the equilibrium
%state, except for special values of $\Gamma$. 
together with thermodynamic equilibrium,
%\subsection{Thermal equilibrium}
%In equilibrium the energy equation implies
\begin{align}\label{thermal_eq}
%  \mathcal{H}_\mathrm{visc} + \mathcal{H}_\mathrm{ext} = \Lambda.   
(q\Omega)^2\rho\nu + \mathcal{H}_\mathrm{ext} = \Lambda,
\end{align}
where the first term represents viscous heating. For the viscous
problem, $\nu\neq0$, we may set $\mathcal{H}_\mathrm{ext}=0$. However,
if we wish to neglect viscosity but include cooling, we must invoke
$\mathcal{H}_\mathrm{ext}\neq0$ to define an equilibrium state.  
To proceed further, we separately describe the two cooling models
considered in this work. 
%It is possible to have both models present - but too complicated

\subsubsection{Beta cooling}
In this case the energy loss per unit volume is specified 
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
\end{align}
where 
\begin{align}
  \tcool = \beta\Omega^{-1} 
\end{align}
is the cooling timescale with $\beta$ a constant input parameter, and
\begin{align}
  \tirr = \theta T_\mathrm{eq}
\end{align}
is a reference temperature field, here parameterized as a constant fraction
$\theta$ of the equilibrium temperature $T_\mathrm{eq}$. We will refer
to $\theta$ as the `irradiation level'.  
%This cooling function restores 
%the disk temperature to $\tirr$ over a time-scale $\tcool$.

For models with beta cooling we assume an equilibrium polytropic relation
\begin{align} 
  P  =
%  \frac{c_{s0}^2\rho_0}{\Gamma}\left(\frac{\rho}{\rho_0}\right)^\Gamma,
  %  \equiv \frac{c_s^2(z)}{\Gamma}\rho,
c_{s0}^2\rho_0\left(\frac{\rho}{\rho_0}\right)^\Gamma,
\end{align}
where $\Gamma$ is the constant polytropic index and $\rho_0 =
\rho(0)$. The vertical
structure is first obtained from Eq. \ref{vert_eq1}---\ref{vert_eq2}
then inserted into Eq. \ref{thermal_eq} to infer the required
viscosity profile,

\begin{align}\label{alpha_beta_relation}
\alpha(z) = \frac{1}{(\gamma-1)\beta
   q^2}\frac{c_s^2(z)}{c_{s0}^2}\left(1 - \theta\right). 
%  \alpha(z) = \frac{1}{\Gamma(\gamma-1)\beta
%    q^2}\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left(1 - \theta\right). 
\end{align}

 Beta cooling is widely applied in 2D and 3D
numerical simulations of self-gravitating disks \citep{gammie01,
  rice05,baehr15}. It permits numerical experiments to be carried out in a
controlled manner as a function of the cooling time $\beta$. 
We remark that for the linear problem  
any explicit cooling function $\Lambda=\Lambda(\rho,T)$ can be
mapped to that obtained from the standard function in Eq.
\ref{beta_cool}. We do this in \S\ref{2dppd} where we adopt a more
realistic beta cooling prescription for 2D PPDs.

>>>>

\section{Basic equations}\label{basic}
We consider a 3D, self-gravitating,
viscous disk with heating and cooling. We use the shearing box framework to
study a small patch of the disk \citep{goldreich65}. The local frame
co-rotates with a fiducial point in the unperturbed disk at angular
frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ corresponds
to the radial, azimuthal and verical directions in the global disk. 
The fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) = 0, \\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi - 2\Omega\hat{\bm{z}}\times\bm{v}\notag\\ & 
  \phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =}
  +2\Omega^2qx\hat{\bm{x}} -\Omega_z^2z\hat{\bm{z}} 
   + \frac{1}{\rho}\nabla\cdot\bm{T},\label{momentum_eq}\\
  & \frac{\p E}{\p t} + \nabla\cdot\left(E\bm{v}\right) = - P
  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} - \Lambda +
  \mathcal{H}_\mathrm{ext} ,\label{energy_eq}
\end{align} 
where $\rho$ is the density field and $\bm{v} = (v_x, v_y, v_z)$ is
the velocity field.% $q$ is the shear parameter and $\Omega_z$ is the
%local vertical oscillation frequency. 
We assume an ideal gas so that the pressure $P$ and thermal energy
density $E$ are related by  
\begin{align}
  P = (\gamma-1)E = \mathcal{R}\rho T, 
\end{align}
where $\mathcal{R}$ is the gas constant and $T$ is the temperature. We
use $\gamma=1.4$ unless otherwise stated. The gas 
gravitatational potential $\Phi$ is given via the Poisson equation,  
\begin{align}
 \nabla^2\Phi = 4 \pi G \rho. 
\end{align}
%where $G$ is the gravitational constant.

@@

In the energy equation (Eq. \ref{energy_eq}) the source terms 
$\mathcal{H}_\mathrm{visc}$ and $\Lambda$
represent viscous heating and time-dependent cooling, 
respectively, and $\mathcal{H}_\mathrm{ext}$ represents any
time-independent heat source/sinks. We normally set
$\mathcal{H}_\mathrm{ext}=~0$ unless otherwise stated.

\subsection{Viscosity and heating}\label{visc_model}
The Cartesian components of the viscous stress tensor $\bm{T}$ is
defined by 
\begin{align}
  T_{ij} \equiv \rho \left[\nu \left(\p_j v_i + \p_i v_j\right) +
    \left(\nu_b-\frac{2}{3}\nu\right)\delta_{ij}\nabla\cdot\bm{v}\right], 
\end{align}
%bulk visc ignored 
where $\rho\nu$ is the shear viscosity. We also include a bulk
viscosity $\rho\nu_b$ for completeness, but will neglect it in
numerical calculations. The associated viscous heating given by 
\begin{align}
  \mathcal{H}_\mathrm{visc} \equiv \left(\p_jv_i\right)T_{ij}, 
\end{align}
where summation over repeated indices is implied.

We adopt the viscosity law  
\begin{align}\label{visc_law}
  \nu = \alpha
  \frac{c_{s0}^2}{\Omega}\left(\frac{\rho}{\rho_\mathrm{eq}}\right)^\mu\left(\frac{P}{P_\mathrm{eq}}\right)^\lambda,             
\end{align}
where subscript `eq' denotes the equilibrium state and  
$c_{s0}^2\equiv P_\mathrm{eq}(z=0)/\rho_\mathrm{eq}(z=0)$. 
The dimensionless viscosity coefficient 
$\alpha=~\alpha(\rho_\mathrm{eq},P_\mathrm{eq})$ characterizes the
magnitude of the shear viscosity \emph{in steady state}. The indices
$\mu,\,\lambda$ are free parameters chosen to model how the viscosity
behaves in the perturbed state.  
We adopt the same prescription for the bulk viscosity but
with $\alpha\to\alpha_b$.

Our numerical calculations use $\mu=-1,\lambda=0$ so 
that $\rho\nu$ is time-indpendent, folling previous 
studies of viscous GI  
\citep{lynden-bell74,hunter83,willerding92,gammie96}.  
This choice also conveniently eliminates viscous over-stability
\citep{schmit95,latter06}, which is unrelated to self-gravity, that
would otherwise contaminate our results. %obscure

While steady-state viscosity values can be determined 
analytically \citep[e.g.][]{rafikov15}, its time-dependent behavior is
not well-explored. We emphasize the choice $\mu=-1,\lambda=0$ is made
to bring out the physical process of interest --- viscous GI. %expect
                                %to 
%be generally present, but will 
%be obscured by viscous over-stability with
%other indices
Interestingly, though, \cite{laughlin96b} have suggested a $\nu \propto 1/\Sigma$
dependence when modeling the evoluting of 2D self-gravitating disks 
as a viscous process. We note that if $\rho\nu$ is constant in time
then increasing the density leads to reduction in the viscosity. 
This is perhaps not inconsistent with numerical simulations of disk
fragmentation which show the internal structure of clumps, which
represent increased density, is laminar \citep{gammie01}.  
{\bf is this true?}

%Furthermore, \cite{paardekooper12} measured
%an increase in the effective viscosity when their gravito-turbulent
%disk models underwent fragmentation (i.e. as $\Sigma$ increases).  
%{\bf note paardekooper 2012: alpha goes down as Sigma goes up}

\subsection{Steady states and cooling models}
We consider equilibrium solutions (here omitting the `eq' subscripts
for simplicity)  
\begin{align}
  \bm{v} &= -q\Omega x \hat{\bm{y}}, \\
  \rho   &= \rho(z),\\
  P      &= P(z) \equiv c_s^2(z)\rho.
\end{align} 
The equilibrium density and pressure fields are obtained by solving
the vertical momentum equation with self-gravity,
\begin{align}
  &\frac{1}{\rho}\frac{dP}{dz} +
  \Omega_z^2z + \frac{d\Phi}{dz} = 0, \label{vert_eq1}\\
 &\frac{d^2\Phi}{dz^2} = 4 \pi G \rho,\label{vert_eq2}
\end{align}
together with thermodynamic equilibrium,
\begin{align}\label{thermal_eq}
(q\Omega)^2\rho\nu + \mathcal{H}_\mathrm{ext} = \Lambda,
\end{align}
where the first term represents viscous heating. For the viscous
problem, $\nu\neq0$, we may set $\mathcal{H}_\mathrm{ext}=0$. However,
if we wish to neglect viscosity but include cooling, we must invoke
$\mathcal{H}_\mathrm{ext}\neq0$ to define an equilibrium state.  
To proceed further, we separately describe the two cooling models
considered in this work. 
%It is possible to have both models present - but too complicated

\subsubsection{Beta cooling}
In this case the energy loss per unit volume is specified 
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}. 
\end{align}
Recall from Eq. \ref{beta_def} that 
%\begin{align}
  $\tcool = \beta\Omega^{-1}$ is 
%\end{align}
the cooling timescale with $\beta$ a constant input parameter. Here, 
\begin{align}
  \tirr = \theta T_\mathrm{eq}
\end{align}
is a reference temperature field parameterized as a constant 
fraction $\theta$ of the equilibrium temperature
$T_\mathrm{eq}$. Physically, $\tirr$ is the floor temperature set by,
for example, stellar irradiation. We will thus refer to $\theta$ as
the `irradiation level'. %stellar illum or env rad

We will formulate the linear problem with beta cooling using the
functional form for $\Lambda$ as given by
Eq. \ref{beta_cool}. However, any explicit cooling function, say
$\Lambda=\Lambda_1(\rho,T)$, can be mapped to that obtained from
Eq. \ref{beta_cool} by equating their linearized forms, i.e. $\delta
\Lambda \equiv \delta \Lambda_1$, which then defines the $\beta$ and
$\theta$ parameters to be used in the framework we develop. This is
done in \S\ref{2dppd} where we adopt a more realistic beta cooling 
prescription for 2D PPDs. In this case, $\theta$ may or may not
represent a physical irradiation.

For models with beta cooling given by Eq. \ref{beta_cool} we assume an 
equilibrium polytropic relation 
\begin{align} 
  P  =
%  \frac{c_{s0}^2\rho_0}{\Gamma}\left(\frac{\rho}{\rho_0}\right)^\Gamma,
  %  \equiv \frac{c_s^2(z)}{\Gamma}\rho,
c_{s0}^2\rho_0\left(\frac{\rho}{\rho_0}\right)^\Gamma,
\end{align}
where $\Gamma$ is the constant polytropic index and $\rho_0 =
\rho(z=0)$ is the equilibrium midplane density. 
The vertical structure is first obtained from
Eq. \ref{vert_eq1}---\ref{vert_eq2} then inserted into
Eq. \ref{thermal_eq} to infer the required viscosity profile for
thermal equilibrium,   
\begin{align}\label{alpha_beta_relation}
\alpha(z) = \frac{1}{(\gamma-1)\beta
   q^2}\frac{c_s^2(z)}{c_{s0}^2}\left(1 - \theta\right). 
%  \alpha(z) = \frac{1}{\Gamma(\gamma-1)\beta
%    q^2}\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left(1 - \theta\right). 
\end{align}

Beta cooling is widely applied in 2D and 3D
numerical simulations of self-gravitating disks \citep{gammie01,
  rice05,baehr15}. It permits numerical experiments to be carried out in a
controlled manner as a function of the cooling time $\beta$. An
adiabatic disk corresponds to $\beta\to \infty$. The
physical meaning of the limit $\beta\to0$ depends on $\theta$, as
discussed in \S\ref{theta0}---\ref{theta1}.

### 2016-01-23 paper/paper3d.tex
model version 2

\begin{abstract}
Low mass, self-gravitating accretion disks admit quasi-steady,
`gravito-turbulent' states in which cooling balances turbulent
heating. However, numerical simulations show that gravito-turbulence
cannot be sustained beyond dynamical timescales when the cooling rate
or turbulent viscosity is too large, resulting in disk fragmentation.        
% `alternative' or `physical' here?
%quantify or develop
We motivate and quantify an alternative interpretation of disk
fragmentation as the inability to maintain gravito-turbulence due to  
secondary gravitational instabilities driven by:  
\begin{inparaenum}[1)] 
\item 
  cooling, which reduces pressure support; and/or
\item 
  viscosity, which reduces rotational support. 
\end{inparaenum}
We analyze the gravitational stability of accretion disks 
including these non-ideal effects explicitly.    
We derive a general dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal heating,
external irradiation, and cooling. We  
consider parameterized cooling functions in two and three dimensions,
as well as full radiative diffusion in 3D. We show that 
cooling/viscosity can destabilize the disk even when the classic
Toomre stability condition is satisfied; and generally there is
no critical cooling rate/viscosity below which the disk is formally
stable. %there is no special/precise cooling/visc that marks the
        %stable/unstable boundary
We apply linear theory to 
global models of protoplanetary disks (PPDs) subject to
gravito-turbulence modeled as an effective 
viscosity and cooling via dust opacity. 
%We find the sole effect of
%cooling is insufficient to 
%destabilize a realistic PPD. 
We find that viscosity renders the PPD beyond $\sim 60$AU dynamically
unstable on radial %this is for Mdot = 1d-6  
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. 
We suggest turbulent stresses can play an active role in realistic disk 
fragmentation by removing rotational stabilization against 
self-gravity. {\bf shorten abstract more?}
\end{abstract}

@@

\subsection{Cooling-driven gravitational instability}\label{cool_gi}
Cooling reduces pressure support against self-gravity. The
cooling time  
$\tcool$ is the time-scale over which the disk temperature $T$ is relaxed
to some floor value $T_0$, which may be zero. That is, $\p_t T =
-(T-T_0)/\tcool$. It is often written as 
\begin{align} 
  \tcool = \beta \Omega^{-1},
\end{align}
where $\beta$ is the corresponding dimensionless cooling time. This
type of cooling function, first applied by \cite{gammie01}, is a
simple parameterization that allows the full range of thermodynamic
response to be explored. 
{\bf physical origin of cooling in PPDs}

@@

In this work, we use viscosity to model such a 
physical effect due to turbulence, but we note that viscous 
GI does not assume any origin for the viscosity. We will extend theory
of viscous GI to include an energy equation  with viscous heating,
irradiation, explicit cooling, as well as three-dimensionality. This
allows us to consider viscous GI in more realistic PPD models.

@@

Indeed, we will find that the region where a gravito-turbulent PPD is
dynamically unstable agrees with the current consensus on  where, and
the conditions under which, a PPD is expected to fragment based on
Eq. \ref{frag_cond}, with typical critical values cited in the
literature \citep[e.g. $\beta_c\sim
  3,\,\alpha_c\sim0.1$;][]{gammie01,rice05}.

@@

The basic equations, disk equilibria, and the two cooling models  
considered --- parametrized `beta' cooling and radiative diffusion, 
are given in \S\ref{basic}. 
The linear stability problem is defined in
\S\ref{linear}. We present results with beta cooling in
\S\ref{result_2d} and \S\ref{3ddisk} for 2D and 3D disks,
respectively. %key result: generalized dispersion relation 
In \S\ref{2dppd} we consider PPDs with realistic viscosity/cooling   
models, including radiative diffusion in 3D, and determine where such
disks are gravitationally unstable. We summarize our results in
\S\ref{summary} with a discussion of how our models may aid the 
physical understanding of fragmentation in realistic PPDs.

>>>>

\begin{abstract}
Low mass, self-gravitating accretion disks admit quasi-steady,
`gravito-turbulent' states in which cooling balances turbulent
heating. However, numerical simulations show that gravito-turbulence
cannot be sustained beyond dynamical timescales when the cooling rate
or turbulent viscosity is too large, resulting in disk fragmentation.        
% `alternative' or `physical' here?
%quantify or develop
We motivate and quantify an alternative interpretation of disk
fragmentation as the inability to maintain gravito-turbulence due to  
secondary gravitational instabilities driven by:  
\begin{inparaenum}[1)] 
\item 
  cooling, which reduces pressure support; and/or
\item 
  viscosity, which reduces rotational support. 
\end{inparaenum}
We analyze the gravitational stability of accretion disks 
including these non-ideal effects explicitly.    
We derive a general dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal heating,
external irradiation, and cooling. We  
consider parameterized cooling functions in two and three dimensions,
as well as full radiative diffusion in 3D. We show that 
cooling/viscosity can destabilize the disk even when the classic
Toomre stability condition is satisfied; and generally there is
no critical cooling rate/viscosity below which the disk is formally
stable. %there is no special/precise cooling/visc that marks the
        %stable/unstable boundary
We apply linear theory to 
global models of protoplanetary disks (PPDs) subject to
gravito-turbulence modeled as an effective 
viscosity and cooling via dust opacity. 
%We find the sole effect of
%cooling is insufficient to 
%destabilize a realistic PPD. 
We find that viscosity renders the PPD beyond $\sim 60$AU dynamically
unstable on radial %this is for Mdot = 1d-6  
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. 
We suggest turbulent stresses can play an active role in realistic disk 
fragmentation by removing rotational stabilization against 
self-gravity. {\bf shorten abstract?}
\end{abstract}

@@

\subsection{Cooling-driven gravitational instability}\label{cool_gi}
Cooling reduces pressure support against self-gravity. The
cooling time  
$\tcool$ is the time-scale over which the disk temperature $T$ is relaxed
to some floor value $T_0$, which may be zero. That is, $\p_t T =
-(T-T_0)/\tcool$. It is often written as 
\begin{align} \label{beta_def}
  \tcool = \beta \Omega^{-1},
\end{align}
where $\beta$ is the corresponding dimensionless cooling time. This
type of cooling function, first applied by \cite{gammie01}, is a
simple parameterization that allows the full range of thermodynamic
response to be explored. 
{\bf physical origin of cooling in PPDs}

@@

In this work, we use viscosity to model such a 
physical effect due to turbulence, but we note that viscous 
GI does not assume any origin for the viscosity. We will generalized
the theory of viscous GI to include an energy equation  with viscous
heating, irradiation, explicit cooling, as well as
three-dimensionality (3D). This 
allows us to consider viscous GI in more realistic PPD models.

@@

Indeed, we find that the region where a gravito-turbulent PPD is
dynamically unstable agrees with the current consensus on  where, and
the conditions under which, a PPD is expected to fragment based on
Eq. \ref{frag_cond}, with typical critical values cited in the
literature \citep[e.g. $\beta_c\sim
  3,\,\alpha_c\sim0.1$;][]{gammie01,rice05}.

@@

The basic equations, disk equilibria, cooling and viscosity models are 
given in \S\ref{basic}. The linear stability problem is defined in 
\S\ref{linear}. We present results with parameterized `beta' cooling in 
\S\ref{result_2d} and \S\ref{3ddisk} for 2D and 3D disks,
respectively. %key result: generalized dispersion relation 
In \S\ref{2dppd} we consider PPDs with realistic viscosity/cooling   
models, including radiative diffusion in 3D, and determine where such
disks are gravitationally unstable. We summarize our results in
\S\ref{summary} with a discussion of how our models may aid the 
physical understanding of fragmentation in realistic PPDs.

### 2016-01-23 paper/results2d.tex
reg update
model version 2

Eq. \ref{inviscid} is a cubic equation in $S$. The 
Routh-Hurwitz criteria imply that stability is ensured if 
\begin{align}
  \gamma > \theta \quad \text{and} \quad 
  Q > \frac{1}{\sqrt{2\theta(2-q)}} 
\end{align}
are both satisfied. Thus, whether or not the system is stable is
independent of the cooling rate $\beta$ (provided it is non-zero and
finite). The system becomes more unstable with decreasing $\theta$.

@@

\subsubsection{$\theta = 0$}  
%Beta cooling with $\theta=0$ is commonly employed in 
%numerical simulations of self-gravitating disks \citep{gammie01}. 
For beta cooling with $\theta=0$, the disk is unconditionally
unstable for finite $Q$. The most unstable wavelength decreases with
decreasing $\beta$. The $\beta\to0$ limit corresponds to a
pressureless disk (not merely isothermal).

@@

\subsubsection{$\theta = 1$}
Beta cooling with $\theta=1$ corresponds to 
`thermal-relaxation' in a strictly inviscid disk: the temperature is restored to its initial
value over the cooling time \citep{lin15,mohandas15}. In this case 
$\beta\to 0$ corresponds to an isothermal disk and $\beta
\to \infty$ corresponds to an adiabatic disk. (These two limits are
equivalent for $\gamma=1$.) The most unstable wavenumber and growth
rates vary weakly with $\beta$.

>>>>

Eq. \ref{inviscid} is a cubic equation in $S$. The 
Routh-Hurwitz criteria imply that stability is ensured if 
\begin{align}\label{stable_condition}
  \gamma > \theta \quad \text{and} \quad 
  Q > \frac{1}{\sqrt{2\theta(2-q)}} 
\end{align}
are both satisfied. Thus, whether or not the system is stable is
independent of the cooling rate $\beta$ (provided it is non-zero and
finite). The system becomes more unstable with decreasing $\theta$.

@@

\subsubsection{$\theta = 0$}\label{theta0}
%Beta cooling with $\theta=0$ is commonly employed in 
%numerical simulations of self-gravitating disks \citep{gammie01}. 
For beta cooling with $\theta=0$, the disk is unconditionally
unstable for finite $Q$. The most unstable wavelength decreases with
decreasing $\beta$. The $\beta\to0$ limit corresponds to a
pressureless disk (not merely isothermal).

@@

\subsubsection{$\theta = 1$}\label{theta1}
Beta cooling with $\theta=1$ corresponds to 
`thermal-relaxation' in a strictly inviscid disk: the temperature is restored to its initial
value over the cooling time \citep{lin15,mohandas15}. In this case 
$\beta\to 0$ corresponds to an isothermal disk and $\beta
\to \infty$ corresponds to an adiabatic disk. (These two limits are
equivalent for $\gamma=1$.) The most unstable wavenumber and growth
rates vary weakly with $\beta$.
